Pythran: Bringing OpenMP to a Python Subset

### 2013-02-07 doc/papers/europar2013/paper.tex
Take into account F. Dagnat's advices.

% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
% max 12 pages
\documentclass{llncs}
%
\usepackage{listings}
\usepackage{url}
\usepackage{todo}
\newcommand\see{\emph{cf.\ }}

@@

\lstset{language=python,morekeywords={cimport, cdef, with}}

\begin{document}
%
\title{Pythran: Bringing OpenMP to a Python Subset}

@@

In the context of interpreted language, \cite{dongara2007} defines a computation
as \emph{hybrid} when part of the code is interpreted, and part of it is
executed natively.

@@

An opposite approach consists in using the host language ---herein Python---
to describe both parts of the system, and to let an automated tool perform the
translation to native code of a part of the application, generally the
computation-intensive one where parallelism has been expressed in some ways.
Thus, developers not familiar with lower level languages or not eager to invest
the additional development time can still benefit from a fair performance boost.
This approach has been subject to many studies that can be classified based on
their backward-compatibity with the host language.

@@

The core issue of these approaches is that they imply to modify, in a more or
less trivial way, the original code. They require to invest in a new language,
and the long-term preservation of this investment is not ensured. Moreover,
there is no falling back if the code where to be deployed in an environment
where the parallelizing tool/module is not available. For instance, once a code
has been ported to PyCUDA and transformed into CUDA code embedded in Python
strings, there is no way back and the developer has to maintain two versions of
the algorithm.

@@

Most backward-compatible approaches also require to modify the input program,
but they do not extend the Python language, rather restrict it. As a
consequence, they remain compatible with the original language and do not suffer
from the drawbacks of the previous approach.

@@

The core directive to handle data parallelism is the \texttt{for} directive that
distribute the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space must be
described by a random access iterator with a total order, i.e.\ it is possible
to increase the iterator by any integer and to decide which iterator is the
greatest among two. The integer used as loop indices in Fortran and C satisfy
this conditions, so do C++ iterators with the \texttt{random\_access\_tag}
trait. But a Python iterator only advance by a step of one until it is
exhausted, in which case it raises an exception. They only have forward iterator
behavior.

The extension of Python iterator to random access iterators is direct for the
standard containers: \texttt{list}~\footnote{Python \texttt{list}s behave as C++
\texttt{vector}s.}, \texttt{set} or \texttt{dict}.

Python also generalizes the concept of iterators through generators, which 
behave like forward iterators. The simplest one, \texttt{xrange(\emph{start},
\emph{stop}, \emph{step})}, successively yields value starting from
\texttt\emph{start} to \texttt\emph{stop} by a step of \texttt\emph{step}. It is
easily extended to support random access, but it generally does not make
sense to use a generator as a loop iterator, as the relation between two
random states of the iterator may be of an arbitrary complexity.

A particular class of generator behaves like adaptor, in the sense that they
apply a particular expression on each value of the iterator. It is the case of
the \texttt{enumerate} builtin, that yields each element of the enumerated
iterator associated with its index, and of generator expressions such as
\texttt{(x*x for x in l)}. These generators are random iterators only if their
input iterator is a random iterator itself.

@@

@@

The OpenMP runtime library makes it possible to retrieve the active thread id,
whether dynamic scheduling is enabled and so on. All the functions are declared
in the \texttt{<omp.h>} header, and have a default behavior when OpenMP is not
activated.

@@

A validation suite for OpenMP has been proposed in~\cite{wang2012} for C and
Fortran. We have ported it to Python, and we also extended it to validate the
corner cases specific to Python described in this section. An example of such
test case is given in Listing~\ref{lst:openmp-validation}.

@@

Pythran is subset of Python focused on scientific computing. It is implicitly
statically typed and support most Python constructs except those that involve
introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
\texttt{eval}). A few standard module are supported in addition to the core
language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
partially supported. The associated compiler turns Pythran code, eventually
annotated with OpenMP and exported function type annotation, into C++ code. Its
processing is summarized in Figure~\ref{fig:pythran-compiler} and relies on type
inference to remove most dynamic behavior, combined with a high-level runtime
library, \texttt{pythonic}, to allow a one-to-one mapping between Python and C++
constructs. The purpose of this section is not to dig into the internal of the
compiler, but rather to focus on the impact of the parallelism layer, especially
on the runtime library.

@@

The standard library of C++11, unlike its C++03 version, is thread-safe. Pythran
runtime is based on the STL and provides the same guaranty, so that the data
dependences in the original code remains explicit.

A critical point of the design of the \texttt{pythonic} runtime is memory
management, the very same aspect that lead to the use of a GIL in CPython.
Memory management is implemented in Shedskin through the general-purpose Boehm
garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
object inside parallel regions, thus making memory management explicit for these
parts.

@@

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. As it limits parallelism, reducing the
number of copies becomes an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing still imply copies. To avoid a reference increment, one can pass
parameters by reference. Using an out-of-scope interprocedural memory effect
analysis, Pythran determines for each argument whether it must be passed using
reference or const reference to prevent this extra cost.

@@

To validate the approach proposed in this paper, we first performed a simple
experiment. Starting from the C code of a small geomatics application, Hyantes,
we successively parallelized it with OpenMP, turned it into Python with a
Pythran compatible kernel and parallelized the Python version using the approach
described in this paper. We also measurer the SLOC for the two versions.
Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It non
only emphasises the use of a high level language prototyping, but also shows
that it is possible to turn this prototype into reasonably efficient code
through Pythran. It is also possible to prototype the parallel version while
remaining at the Python level.

@@

    \centering
    \begin{tabular}{|l|l|l|l|}
        \hline
        Language & SLOC & wall time (s) [gcc] & wall time (s) [clang]\\
        \hline
        C       & 102   & 25 & 22 \\
        Python  & 30    & 676 & --\\
        Pythran & 30    & 38 &  22 \\
        Pythran+OMP    & 30    & 11 & no OpenMP support\\
        \hline
    \end{tabular}

@@

We then compare Pythran with Cython. The approach share some similarities: they
both generate code written in a rather low level language, with OpenMP
directives. However, the inputs differ as Cython requires more typing
information that Pythran to generate efficient code, and Cython is not
backward-compatible with Python. They both translate explicit fine-grained
parallelism through code annotations or specific python constructs,
respectively.

Pythran exposes the full OpenMP interface to the user, thus enabling both data
and task parallelism, as described in Section~\ref{sec:python-openmp}. This is
not the case in Cython, where:
%
\begin{itemize}
    \item Only loops can be made parallel, through the use of a dedicated
        intrinsic \texttt{prange} as loop iterator.

@@

    \item As Cython allows the mix of Python and Cython code, the user is
        responsible from releasing the GIL.

    \item It is impossible to use a function imported from a Python module in a
        parallel region, while it is still possible to use native C functions.

\end{itemize}

Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrate the
difference between the two and illustrates the intrusive behavior of Cython.

@@

This paper studies the relevancy of adding OpenMP annotations to Python code. It
shows that providing minor semantic adaptations, these annotations make it
possible for Python code to benefit from multi-core performance boost while
retaining backward compatibility and without worrying about the Global
Interpreter Lock.

>>>>

% to submit via https://www.easychair.org/account/signin.cgi?conf=europar2013
% max 12 pages
\documentclass{llncs}
%
\usepackage{listings}
\usepackage{url}
\usepackage{todo}
\usepackage{xcolor}
\newcommand\see{\emph{cf.\ }}

@@

\lstset{language=python,morekeywords={cimport, cdef, with},
basicstyle=\ttfamily, keywordstyle=\bf, stringstyle=\color{gray}}

\begin{document}
% update listing numbering
\renewcommand{\thelstlisting}{\arabic{lstlisting}}
%
\title{Pythran: Bringing OpenMP to a Python Subset}

@@

In the context of interpreted languages, \cite{dongara2007} defines a computation
as \emph{hybrid} when part of the code is interpreted, and part of it is
executed natively.

@@

An opposite approach consists in using the host language ---herein Python---
to describe both parts of the system, and to let an automated tool perform the
translation to native code of a part of the application, generally the
computation-intensive one where parallelism has been expressed in some ways.
Thus, developers not familiar with lower level languages or not eager to invest
the additional development time can still benefit from a fair performance boost.
This approach is the subject of many studies that can be classified based on
their compatibility with the host language.

@@

The core issue of these approaches is that they imply to modify, in a more or
less complex way, the original code. They require to invest in a new language,
and the long-term preservation of this investment is not ensured. Moreover,
there is no falling back if the code where to be deployed in an environment
where the parallelizing tool/module is not available. For instance, once a code
has been ported to PyCUDA and transformed into CUDA code embedded in Python
strings, there is no way back and the developer has to maintain two versions of
the algorithm.

@@

Most backward-compatible approaches also require to modify the input program,
but they do not extend the Python language, rather restrict it. As a
consequence, they remain compatible with the original language and do not suffer
from the drawbacks of the previous approach. They also keep the benefit of
existing tools associated to the language.

@@

The core directive to handle data parallelism is the \texttt{for} directive that
distributes the iteration space of the associated loop among the existing
threads. To be compatible with OpenMP, the loop iteration space must be
described by a random access iterator with a total order. Integers used as
loop indices in Fortran and C satisfy this conditions, as well as C++ iterators with
the \texttt{random\_access\_tag} trait. But a Python iterator only advances by a
step of one until it is exhausted, in which case it raises an exception: it
behaves as a forward iterator. To be compatible with OpenMP, these iterators
must be turned into random access iterators.

The extension of Python iterator to random access iterators is direct for the
standard containers: \texttt{list}~\footnote{Python \texttt{list}s behave as C++
\texttt{vector}s.}, \texttt{set} or \texttt{dict}. Other iterators require more
care.

Generators, Python objects that behave like iterators, are commonly used in
Python. The simplest one, \texttt{xrange(\emph{start}, \emph{stop},
\emph{step})}, successively yields value starting from \texttt\emph{start} to
\texttt\emph{stop} by a step of \texttt\emph{step}. It is easily extended to
support random access, but it generally does not make sense to use a generator
as a loop iterator, as the relation between two random states of the iterator
may be of an arbitrary complexity.

Generator expressions are generators whose content is built from an other
iterator. For instance \texttt{(x*x for x in l)} successively yields the square
of each element in \texttt{l}. They behave like adaptors: they apply a
particular expression on each value of the iterator. It is also the case of the
\texttt{enumerate} builtin, that yields each element of the enumerated iterator
associated with its index. These generators are random iterators only if their
input iterator is a random iterator itself.

@@

Pythran runtime is aware of these three kind of iterators: random access
iterators for standard containers, adaptor iterators for generator expressions
and forward iterators for generators.

@@

OpenMP provides a small runtime library that, for instance, makes it possible to
retrieve the active thread id. All the functions are declared in the
\texttt{<omp.h>} header, and have a default behavior when OpenMP is not
activated.

@@

A validation suite for OpenMP is proposed in~\cite{wang2012} for C and Fortran.
We ported it to Python, and we also extended it to validate the corner cases
specific to Python described in this section. A typical test case is given in
Listing~\ref{lst:openmp-validation}.

@@

Pythran is a subset of Python focused on scientific computing. It is implicitly
statically typed and support most Python constructs except those that involve
introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
\texttt{eval}). A few standard module are supported in addition to the core
language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
currently only partially supported. The associated compiler turns Pythran code,
possibly annotated with OpenMP and a few function type annotations, into C++
code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
relies on type inference to remove most dynamic behavior, combined with a
high-level runtime library, \texttt{pythonic++}, to allow a one-to-one mapping
between Python and C++ constructs. The purpose of this section is not to dig
into the internal of the compiler, but rather to focus on the impact of the
parallelism layer, especially on the runtime library.

@@

The standard library of C++11, unlike its C++03 version, is thread-safe. Pythran
runtime is based on the STL and provides the same guaranty, so that the data
dependencies in the original code remains explicit.

A critical point of the design of the \texttt{pythonic++} runtime is memory
management, the very same aspect that led to the use of a GIL in CPython.
Memory management is implemented in Shedskin through the general-purpose Boehm
garbage collector~\cite{boehm1991}, and Cython forbids usage of Python-managed
object inside parallel regions, thus making memory management explicit for these
parts.

@@

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. As it limits parallelism, reducing the
number of copies becomes an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing still implies copies. To avoid a reference increment, one can pass
parameters by reference. Using an out-of-scope interprocedural memory effect
analysis, Pythran determines for each argument whether it must be passed using
reference or const reference to prevent this extra cost.

@@

To validate the approach proposed in this paper, we first performed a simple
experiment. Starting from the C code of a small geomatics application, Hyantes,
we successively parallelized it with OpenMP, turned it into Python with a
Pythran compatible kernel and parallelized the Python version using the approach
described in this paper. We also measure the SLOC of the two versions.
Table~\ref{tbl:hyantes} summarizes the results of the experimentation made on a
laptop with 4 non hyperthreaded cores using \texttt{-O2} compiler flag. It not
only emphasises the use of a high level language prototyping, but also shows
that it is possible to turn this prototype into reasonably efficient code
through Pythran. It is also possible to prototype the parallel version while
remaining at the Python level.

@@

    \centering
    \begin{tabular}{|l|l|l|l|}
        \hline
        Language & SLOC & wall time (s) [gcc] & wall time (s) [clang]\\
        \hline
        C       & 102   & 25 & 22 \\
        Python  & 30    & 676 & --\\
        Pythran & 30    & 38 &  22 \\
        Pythran+OMP    & 31    & 11 & no OpenMP support\\
        \hline
    \end{tabular}

@@

We then compare Pythran with Cython. The approach share some similarities: they
both generate code written in a lower level language, with OpenMP
directives. However, the inputs differ as Cython requires more typing
information that Pythran to generate efficient code, and Cython is not
backward-compatible with Python. They both translate explicit fine-grained
parallelism through code annotations or specific python constructs,
respectively.

Pythran exposes the full OpenMP interface to the user, thus enabling both data
and task parallelism, as described in Section~\ref{sec:python-openmp}. This is
not the case in Cython, where:
%
\begin{itemize}

    \item Only loops can be made parallel, using a new \texttt{prange} intrinsic
        as iterator.

@@

    \item The user is responsible from releasing the GIL inside parallel
        regions.

    \item It is impossible to use a function imported from a Python module in a
        parallel region, but it is still possible to use native C functions.

\end{itemize}
%
Listings~\ref{lst:cython-sample} and~\ref{lst:pythran-sample} illustrate the
difference between the two and illustrates the intrusive behavior of Cython.

@@

This paper studies the addition  of OpenMP annotations to Python. It shows that
providing minor semantic adaptations, these annotations make it possible for
Python code to benefit from multi-cores while retaining backward compatibility
and without worrying about the Global Interpreter Lock.
